Generalized spaces for constructive algebra

### 2020-04-09 paper-herrsching2019/main.tex
Cleaner macros

\usepackage{hyperref}

@@

\newcommand{\A}{\mathcal{A}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\C}{\mathcal{C}}
\newcommand{\M}{\mathcal{M}}
\renewcommand{\AA}{\mathbb{A}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\G}{\mathcal{G}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\GG}{\mathbb{G}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\K}{\mathcal{K}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\PP}{\mathbb{P}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\renewcommand{\P}{\mathcal{P}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\GL}{\mathrm{GL}}
\newcommand{\ab}{\mathrm{ab}}
\newcommand{\Zar}{\mathrm{Zar}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\newcommand{\Sch}{\mathrm{Sch}}
\newcommand{\Aff}{\mathrm{Aff}}
\newcommand{\Ring}{\mathrm{Ring}}
\newcommand{\LocRing}{\mathrm{LocRing}}
\newcommand{\LRS}{\mathrm{LRS}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\Max}{\mathrm{Max}}
\newcommand{\Gal}{\mathrm{Gal}}
\newcommand{\Rad}{\mathrm{Rad}}
\newcommand{\Idl}{\mathrm{Idl}}
\newcommand{\lra}{\longrightarrow}
\newcommand{\RelSpec}{\operatorname{Spec}}
\renewcommand{\_}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\speak}[1]{\ulcorner\text{\textnormal{#1}}\urcorner}
\newcommand{\ul}[1]{\underline{#1}}
\newcommand{\affl}{\ensuremath{{\ul{\ensuremath{\AA}}^1}}}
\newcommand{\Ll}{\text{iff}}
\newcommand{\inv}{inv.\@}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
\newcommand{\hg}{\mathbin{:}}  % homogeneous coordinates
\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\Pt}{\mathrm{Pt}}
\newcommand{\Loc}{\mathrm{Loc}}
\newcommand{\Top}{\mathrm{Top}}
\newcommand{\ann}{\operatorname{ann}}
\newcommand{\BPIT}{\textsc{bpit}\xspace}
\newcommand{\notnot}{\emph{not not}\xspace}
\newcommand{\negg}{\neg\!\!\!\neg}
\newcommand{\bott}{\bot\!\!\!\!\bot}

@@

\textbf{Definition.} A \emph{topological space} consists of a set~$X$ of points
together with a set~$\O(X) \subseteq \P(X)$ of point sets which are deemed
\emph{open} such that (arbitrary, set-indexed) unions and finite intersections
of open sets are open.
\end{quote}

@@

\paragraph{Geometric theories and sheaf models.} The most sensible solution to
the task
\begin{quote}
Determine the continuous complex-valued functions~$f$ with~$f(z)^2 = z$.
\end{quote}
is not the single set
\[ \{ f : \CC \to \CC \,|\, \text{$f$ is continuous and $f(z)^2 = z$ for all~$z \in \CC$} \}. \]
Rather, since the solvability of this functional equation varies with the
domain on which~$f$ should be defined (having no global solution defined on all
of~$\CC$ but two solutions on small disks not containing the origin), we should
consider the family
\[ (\{ f : U \to \CC \,|\, \text{$f$ is continuous and $f(z)^2 = z$ for all~$z
\in \CC$} \})_{U \in \O(\CC)} \]
of sets as the solution (or perhaps, since no solution was actually given, as
an apt description) of the task.

@@

\paragraph{Random sequences.} \emph{Cantor space} is the space~$2^\NN$ of
infinite binary sequences, equipped with the product topology. A subset~$U
\subseteq 2^\NN$ is open with respect to this topology if and only if for
every~$\alpha \in U$, there is a number~$n \in \NN$ such that
\[ \{ \beta \in 2^\NN \,|\, \forall i < n\_ \alpha(i) = \beta(i) \} \subseteq U. \]

The open subsets of~$2^\NN$ are precisely those subsets which encode
\emph{observable} (or \emph{semidecidable}) properties of binary sequences,
those properties for which a finite record of digits suffices to verify that a
sequence has it (but not necessarily to falsify it). For instance, the sets~$V
= \{ \alpha \in 2^\NN \,|\, \alpha(0) = \alpha(1) = 1 \}$ and~$W = \{
\alpha \in 2^\NN \,|\, \exists n\in\NN\_ \alpha(n) = 1 \}$ are open and the sets
\begin{align*}
  A &\defeq \{ \alpha \in 2^\NN \,|\, \forall n\in\NN\_ \alpha(n) = 0 \} \\
  B &\defeq \Bigl\{ \alpha \in 2^\NN \,\Big|\, \lim_{n \to \infty} \frac{1}{n} \sum_{i=0}^{n-1} \alpha(i) = \frac{1}{2} \Bigr\}
\end{align*}
are not. With respect to the standard Lebesgue measure on~$2^\NN$, the sets~$V$
and~$W$ have measure~$\frac{1}{4}$ and~$1$, respectively.

@@

However, the resulting theory will be trivial for the plain reason that
according to this definition, there are no random sequences: For any
sequence~$\alpha$, the set~$\{ \beta \in 2^\NN \,|\, \exists n \in \NN\_
\alpha(n) \neq \beta(n) \}$ is open, has measure~$1$ and does not
contain~$\alpha$.

@@

The starting point of the theory of locales is the following
observation. The set~$\O(X)$ of open subsets of a topological
space~$X$ forms a partially ordered set which has
\[ \begin{array}{@{}ccc@{}}
\text{arbitrary joins (suprema)}
&\quad\text{and}\quad&
\text{finite meets (infima)}, \\
\bigvee && \wedge
\end{array} \]
and finite meets distribute over arbitrary joins:
\[ U \wedge \bigvee_i V_i = \bigvee_i (U \wedge V_i). \]

@@

\begin{definition}\label{defn:locale}
A \emph{locale}~$X$ is given by a frame~$\O(X)$, the ``frame
of opens of~$X$''. A \emph{morphism~$f : X \to X'$ of locales} (or ``continuous
map of locales'') is a frame homomorphism~$\O(X') \to \O(X)$.\end{definition}

@@

Examples for locales include the following.
\begin{enumerate}
\item Any topological space~$Y$ induces a locale~$L(Y)$ by
setting~$\O(L(Y)) \defeq \O(Y)$. A continuous map~$f : Y \to Y'$ of topological
spaces induces the frame homomorphism~$\O(Y') \to \O(Y),\,U \mapsto f^{-1}[U]$
in the other direction and hence a morphism~$L(Y) \to L(Y')$ of locales in the
same direction.
\item The \emph{one-point locale}~$\pt$ is the locale induced
by the one-point topological space~$\{\star\}$. Its frame of opens is
the powerset of~$\{\star\}$, also known as the set~$\Omega$ of truth
values. Its least element is~$\bot = \emptyset$ and its largest element
is~$\top = \{\star\}$, and potentially not all elements of~$\Omega$ are equal
to one of these two.
\item The locale of surjections~$\NN \twoheadrightarrow \RR$ and the localic
version of the reals of Section~\ref{sect:examples-no-points} are best
constructed as \emph{classifying locales}, a notion to be introduced in
Section~\ref{sect:presenting-frames}.
\item The radical ideals of any ring~$A$ form a frame and hence give rise to a
locale. We will identify this locale in Theorem~\ref{thm:workhorse} to be the
\emph{spectrum} of~$A$ as studied in algebraic geometry.
\end{enumerate}

@@

The underlying frame homomorphism~$\alpha : \O(X) \to \Omega$ of a point~$x$
of~$X$ can be pictured as mapping each open~$U$ of~$X$ to the truth value to which
extent~$x$ belongs to~$U$. Symbolically, we write~$x \inplus U$ iff~$\alpha(U)
= \top$.

The point~$x$ is completely determined by the information to which opens it
belongs, that is by the set {$\{ U \in \O(X) \,|\, \alpha(U) = \top \}$}. This set is a \emph{completely
prime filter}, that is a subset~$\fff \subseteq \O(X)$ which is upward-closed,
closed under finite meets and for which~$\bigvee_i U_i \in \fff$ implies~$U_i
\in \fff$ for some index~$i$; and conversely, any such completely prime filter
gives rise to a point of~$X$ (Exercise~\ref{ex:points-as-filters}).

\begin{definition}A locale~$X$ is \emph{spatial} if and only if its points
suffice to detect the inclusion relation on the opens of~$X$, that is if for any opens~$U, V \in \O(X)$,
if~$x \inplus U \Rightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U
\preceq V$.\end{definition}

An equivalent definition is: A locale~$X$ is \emph{spatial} if and only if its points
suffice to distinguish its opens, that is if for any opens~$U, V \in \O(X)$, if~$x \inplus U
\Leftrightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U = V$.

@@

{\scriptsize\emph{Hint.} Given a completely prime filter~$\fff \subseteq
\O(Y)$, the set~$\bigcup\{ U \in \O(Y) \,|\, U \not\in \fff \}$ is a
meet-irreducible open subset.\par}
\item Assuming the law of excluded middle, show that a topological space is
sober in the sense of Remark~\ref{rem:definition-sober} if and only if for
every meet-irreducible open subset~$W$ there is a unique point~$y \in Y$ such
that~$W = Y \setminus \overline{\{y\}}$.
(This is a classical definition of sobriety.)

@@

\begin{exercise}{Complete metric spaces are sober}
A \emph{metric space} is a set~$Y$ together with a relation~$d \subseteq Y
\times Y \times \QQ_{\geq0}$, written~``$d(x,y) \leq q$'', such that the
following axioms are satisfied.
\begin{align*}
  d(x,y) \leq q \wedge q \leq r &\Longrightarrow d(x,y) \leq r \\
  (\forall r \in \QQ_{> q}\_ d(x,y) \leq r) &\Longrightarrow d(x,y) \leq q \\
  &\mathrel{\phantom{\Longrightarrow}} \exists q \in \QQ_{\geq0}\_ d(x,y) \leq q \\
  &\mathrel{\phantom{\Longrightarrow}} d(x,x) \leq 0 \\
  d(x,y) \leq 0 &\Longrightarrow x = y \\
  d(x,y) \leq q &\Longrightarrow d(y,x) \leq q \\
  d(x,y) \leq q \wedge d(y,z) \leq r &\Longrightarrow d(x,z) \leq q+r
\end{align*}
A \emph{Cauchy process} (or \emph{multi-valued modulated Cauchy sequence}) in a metric
space~$Y$ is a map~$\alpha : \QQ_{>0} \to P(Y)$ such that all the sets~$\alpha(q)$
are inhabited and such that~$d(x,x') \leq q + q'$ for all~$q,q' \in \QQ_{>0}$
and~$x \in \alpha(q), x' \in \alpha(q')$. Such a process \emph{converges
to}~$x_0 \in Y$ iff~$d(x,x_0) \leq q$ for all~$q \in \QQ_{>0}$
and~$x \in \alpha(q)$. A metric space~$Y$ is \emph{complete} iff every Cauchy
process in~$Y$ converges.

@@

\begin{exercise}{The covariant approach to locales}%
Let~$A$ be a frame.
\begin{alphlist}[(c)]
\item Show that~$A$ contains arbitrary (set-indexed) meets (though they are not
required to be preserved by frame homomorphisms), by the impredicative
construction
\[ \bigwedge_{i \in I} U_i = \bigvee\{ V \in A \,|\, \text{$V \preceq U_i$ for all~$i \in I$}
\}. \]
\item What are the arbitrary meets in the frame of open subsets of a
topological space?
\item Show that for any frame homomorphism~$\alpha : A \to A'$ there exists a
monotone map~$\beta : A' \to A$ such that~\emph{$\beta$ is right adjoint
to~$\alpha$}, that is such that
\[ \alpha(U) \preceq V \quad\text{iff}\quad U \preceq \beta(V) \]
for all~$U \in A$ and~$V \in A'$.
\end{alphlist}
{\scriptsize\emph{Note.} Based on the observation in part~(c), a
\emph{covariant approach} to locales can be developed, in which locale maps~$X
\to X'$ are defined to be certain kinds of maps~$\O(X) \to
\O(X')$~\cite{picado-pultr:covariant}.\par}
\end{exercise}

\begin{exercise}{Points as complete prime filters}%
\label{ex:points-as-filters}%
Verify that the points of a locale~$X$ are in canonical one-to-one correspondence
with the completely prime filters of~$\O(X)$.
\end{exercise}

\begin{exercise}{Dense opens}%
\label{ex:dense-opens}%
An open~$U$ of a locale is \emph{dense} iff for any open~$V$, $U \wedge V =
\bot$ implies~$U = \bot$.
\begin{alphlist}[(b)]
\item Let~$M$ be a set and endow it with the discrete topology. Show that an
open~$U \in \O(L(M))$ is dense iff for any element~$x \in M$, $\neg\neg(x \in
U)$.
\item Let~$\varphi$ be a truth value, hence an open of the one-point
locale~$\pt$. Show that~$\varphi$ is dense in~$\pt$ iff~$\neg\neg\varphi$.
\end{alphlist}
\end{exercise}

\begin{exercise}{Characterizing spatial locales}%
Let~$X$ be a locale. Show that the following conditions are equivalent.
\begin{alphlist}[(c)]
\item For any opens~$U, V \in \O(X)$: If~$x \inplus U \Rightarrow y \inplus V$
for all points~$x$ of~$X$, then~$U \preceq V$.
\item For any opens~$U, V \in \O(X)$: If~$x \inplus U \Leftrightarrow y \inplus V$
for all points~$x$ of~$X$, then~$U = V$.
\item There is a topological space~$Y$ and a surjective morphism~$L(Y) \to X$
of locales. (A morphism of locales is \emph{surjective} iff its underlying
frame homomorphism is injective.)
\item The canonical morphism~$L(\Pt(X)) \to X$ is an isomorphism.
\end{alphlist}
\end{exercise}

@@

    \textnormal{double rule for existential quantification} \\\smallskip
    \phantom{a}\hfill
    \Axiom$\varphi \ \fCenter\seq{\vec x, y} \psi$
    \doubleLine
    \UnaryInf$\exists y\?Y\_ \varphi\ \fCenter\seq{\vec x} \psi$
    \DisplayProof
    \makebox[0pt]{\shifttext{\tiny ($y$ not free in~$\psi$)}}
    \hfill\phantom{a}
    \bigskip

    \textnormal{mixed rules (unnecessary in presence of the rule for implication)} \\\smallskip
    \vspace{-0.5em}
    \phantom{a}\hfill
    \AxiomC{$\phantom{\seq{\vec x}}$}
    \UnaryInfC{$(\bigvee_i \varphi_i) \wedge \psi \seq{\vec x} \bigvee_i (\varphi_i \wedge \psi)$}
    \DisplayProof
    \hfill
    \AxiomC{$\phantom{\seq{\vec x}}$}
    \UnaryInfC{$(\exists y\?Y\_ \varphi) \wedge \psi \seq{\vec x} \exists y\?Y\_ (\varphi \wedge \psi)$}
    \DisplayProof
    \makebox[0pt]{\shifttext{\tiny ($y$ none of the~$\vec x$)}}
    \hfill\phantom{a}
    \bigskip

@@

    \textnormal{double rule for universal quantification} \\\smallskip
    \phantom{a}\hfill
    \Axiom$\varphi \ \fCenter\seq{\vec x, y} \psi$
    \doubleLine
    \UnaryInf$\varphi\ \fCenter\seq{\vec x} \forall y\?Y\_ \psi$
    \DisplayProof
    \makebox[0pt]{\shifttext{\tiny ($y$ not free in~$\varphi$)}}
    \hfill\phantom{a}
    \bigskip

@@

By adding the axiom
\begin{align*}
  \top &\seq{x:R} (x = 0) \vee (\exists y\?R\_ xy = 1),
\intertext{we describe \emph{geometric fields}, and by additionally adding for
each natural number~$n \geq 1$ the axiom}
  \top &\seq{a_0:R,\ldots,a_{n-1}:R} \exists x\?R\_ x^n + a_{n-1}x^{n-1} +
  \cdots + a_1x + a_0 = 0,
\end{align*}
we describe \emph{algebraically closed geometric fields}. (This field
condition is called ``geometric'' because it can be put as a geometric sequent.
Other, constructively not equivalent conditions such as ``any nonzero element
is invertible'' or ``any nonunit is zero'' are not geometric sequents.)

@@

A variant of this theory is the theory of inhabited objects, which has the
axiom~$\top \vdash \exists x\?X\_ \top$. Its set-based models are precisely the
inhabited sets.

@@

The Lindenbaum algebra is a frame, with finite meets given by~$[\varphi] \wedge
[\psi] = [\varphi \wedge \psi]$ and set-indexed joins given by~$\bigvee_i
[\varphi_i] = [\bigvee_i \varphi_i]$;\footnote{This short description of the
set-indexed joins picks representatives from each equivalence class and is
hence only sensible in the presence of the axiom of choice. In a
constructive metatheory, we should rather write~$\bigvee_{i \in I} M_i =
[\bigvee_{\varphi \in K} \varphi]$, where~$K = \{ \varphi \,|\, \exists i \in
I\_ \varphi \in M_i \}$.} we can regard the Lindenbaum algebra as the free
frame generated by the nullary relation symbols of~$\TT$ modulo the axioms
of~$\TT$. Our interest in the Lindenbaum algebra is because it gives rise to a
locale:

>>>>

@@

\newcommand{\CCC}{\mathcal{C}}
\newcommand{\OOO}{\mathcal{O}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\QQ}{\mathbb{Q}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\TT}{\mathbb{T}}
\newcommand{\ZZ}{\mathbb{Z}}
\newcommand{\CC}{\mathbb{C}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\ppp}{\mathfrak{p}}
\newcommand{\qqq}{\mathfrak{q}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\fff}{\mathfrak{f}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\defeqv}{\vcentcolon\equiv}
\newcommand{\Sh}{\mathrm{Sh}}
\newcommand{\ab}{\mathrm{ab}}
\newcommand{\op}{\mathrm{op}}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Hom}{\mathrm{Hom}}
\newcommand{\Spec}{\mathrm{Spec}}
\newcommand{\Max}{\mathrm{Max}}
\newcommand{\Gal}{\mathrm{Gal}}
\newcommand{\Rad}{\mathrm{Rad}}
\newcommand{\Idl}{\mathrm{Idl}}
\newcommand{\+}{\mathpunct{.}}
\newcommand{\?}{\,{:}\,}
\newcommand{\seq}[1]{\mathrel{\vdash\!\!\!_{#1}}}
\newcommand{\pt}{\mathrm{pt}}
\newcommand{\Pt}{\mathrm{Pt}}
\newcommand{\Loc}{\mathrm{Loc}}
\newcommand{\Top}{\mathrm{Top}}
\newcommand{\ann}{\operatorname{ann}}
\newcommand{\BPIT}{\textsc{bpit}\xspace}
\newcommand{\notnot}{\emph{not not}\xspace}
\newcommand{\negg}{\neg\!\!\!\neg}
\newcommand{\bott}{\bot\!\!\!\!\bot}
\newcommand{\brak}[1]{{\llbracket{#1}\rrbracket}}

@@

\textbf{Definition.} A \emph{topological space} consists of a set~$X$ of points
together with a set~$\OOO(X) \subseteq \P(X)$ of point sets which are deemed
\emph{open} such that (arbitrary, set-indexed) unions and finite intersections
of open sets are open.
\end{quote}

@@

\paragraph{Geometric theories and sheaf models.} The most sensible solution to
the task
\begin{quote}
Determine the continuous complex-valued functions~$f$ with~$f(z)^2 = z$.
\end{quote}
is not the single set
\[ \{ f : \CC \to \CC \,|\, \text{$f$ is continuous and $f(z)^2 = z$ for all~$z \in \CC$} \}. \]
Rather, since the solvability of this functional equation varies with the
domain on which~$f$ should be defined (having no global solution defined on all
of~$\CC$ but two solutions on small disks not containing the origin), we should
consider the family
\[ (\{ f : U \to \CC \,|\, \text{$f$ is continuous and $f(z)^2 = z$ for all~$z
\in \CC$} \})_{U \in \OOO(\CC)} \]
of sets as the solution (or perhaps, since no solution was actually given, as
an apt description) of the task.

@@

\paragraph{Random sequences.} \emph{Cantor space} is the space~$2^\NN$ of
infinite binary sequences, equipped with the product topology. A subset~$U
\subseteq 2^\NN$ is open with respect to this topology if and only if for
every~$\alpha \in U$, there is a number~$n \in \NN$ such that
\[ \{ \beta \in 2^\NN \,|\, \forall i < n\+ \alpha(i) = \beta(i) \} \subseteq U. \]

The open subsets of~$2^\NN$ are precisely those subsets which encode
\emph{observable} (or \emph{semidecidable}) properties of binary sequences,
those properties for which a finite record of digits suffices to verify that a
sequence has it (but not necessarily to falsify it). For instance, the sets~$V
= \{ \alpha \in 2^\NN \,|\, \alpha(0) = \alpha(1) = 1 \}$ and~$W = \{
\alpha \in 2^\NN \,|\, \exists n\in\NN\+ \alpha(n) = 1 \}$ are open and the sets
\begin{align*}
  A &\defeq \{ \alpha \in 2^\NN \,|\, \forall n\in\NN\+ \alpha(n) = 0 \} \\
  B &\defeq \Bigl\{ \alpha \in 2^\NN \,\Big|\, \lim_{n \to \infty} \frac{1}{n} \sum_{i=0}^{n-1} \alpha(i) = \frac{1}{2} \Bigr\}
\end{align*}
are not. With respect to the standard Lebesgue measure on~$2^\NN$, the sets~$V$
and~$W$ have measure~$\frac{1}{4}$ and~$1$, respectively.

@@

However, the resulting theory will be trivial for the plain reason that
according to this definition, there are no random sequences: For any
sequence~$\alpha$, the set~$\{ \beta \in 2^\NN \,|\, \exists n \in \NN\+
\alpha(n) \neq \beta(n) \}$ is open, has measure~$1$ and does not
contain~$\alpha$.

@@

The starting point of the theory of locales is the following
observation. The set~$\OOO(X)$ of open subsets of a topological
space~$X$ forms a partially ordered set which has
\[ \begin{array}{@{}ccc@{}}
\text{arbitrary joins (suprema)}
&\quad\text{and}\quad&
\text{finite meets (infima)}, \\
\bigvee && \wedge
\end{array} \]
and finite meets distribute over arbitrary joins:
\[ U \wedge \bigvee_i V_i = \bigvee_i (U \wedge V_i). \]

@@

\begin{definition}\label{defn:locale}
A \emph{locale}~$X$ is given by a frame~$\OOO(X)$, the ``frame
of opens of~$X$''. A \emph{morphism~$f : X \to X'$ of locales} (or ``continuous
map of locales'') is a frame homomorphism~$\OOO(X') \to \OOO(X)$.\end{definition}

@@

Examples for locales include the following.
\begin{enumerate}
\item Any topological space~$Y$ induces a locale~$L(Y)$ by
setting~$\OOO(L(Y)) \defeq \OOO(Y)$. A continuous map~$f : Y \to Y'$ of topological
spaces induces the frame homomorphism~$\OOO(Y') \to \OOO(Y),\,U \mapsto f^{-1}[U]$
in the other direction and hence a morphism~$L(Y) \to L(Y')$ of locales in the
same direction.
\item The \emph{one-point locale}~$\pt$ is the locale induced
by the one-point topological space~$\{\star\}$. Its frame of opens is
the powerset of~$\{\star\}$, also known as the set~$\Omega$ of truth
values. Its least element is~$\bot = \emptyset$ and its largest element
is~$\top = \{\star\}$, and potentially not all elements of~$\Omega$ are equal
to one of these two.
\item The locale of surjections~$\NN \twoheadrightarrow \RR$ and the localic
version of the reals of Section~\ref{sect:examples-no-points} are best
constructed as \emph{classifying locales}, a notion to be introduced in
Section~\ref{sect:presenting-frames}.
\item The radical ideals of any ring~$A$ form a frame and hence give rise to a
locale. We will identify this locale in Theorem~\ref{thm:workhorse} to be the
\emph{spectrum} of~$A$ as studied in algebraic geometry.
\end{enumerate}

@@

The underlying frame homomorphism~$\alpha : \OOO(X) \to \Omega$ of a point~$x$
of~$X$ can be pictured as mapping each open~$U$ of~$X$ to the truth value to which
extent~$x$ belongs to~$U$. Symbolically, we write~$x \inplus U$ iff~$\alpha(U)
= \top$.

The point~$x$ is completely determined by the information to which opens it
belongs, that is by the set {$\{ U \in \OOO(X) \,|\, \alpha(U) = \top \}$}. This set is a \emph{completely
prime filter}, that is a subset~$\fff \subseteq \OOO(X)$ which is upward-closed,
closed under finite meets and for which~$\bigvee_i U_i \in \fff$ implies~$U_i
\in \fff$ for some index~$i$; and conversely, any such completely prime filter
gives rise to a point of~$X$ (Exercise~\ref{ex:points-as-filters}).

\begin{definition}A locale~$X$ is \emph{spatial} if and only if its points
suffice to detect the inclusion relation on the opens of~$X$, that is if for any opens~$U, V \in \OOO(X)$,
if~$x \inplus U \Rightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U
\preceq V$.\end{definition}

An equivalent definition is: A locale~$X$ is \emph{spatial} if and only if its points
suffice to distinguish its opens, that is if for any opens~$U, V \in \OOO(X)$, if~$x \inplus U
\Leftrightarrow x \inplus V$ for all points~$x$ of~$X$, then~$U = V$.

@@

{\scriptsize\emph{Hint.} Given a completely prime filter~$\fff \subseteq
\OOO(Y)$, the set~$\bigcup\{ U \in \OOO(Y) \,|\, U \not\in \fff \}$ is a
meet-irreducible open subset.\par}
\item Assuming the law of excluded middle, show that a topological space is
sober in the sense of Remark~\ref{rem:definition-sober} if and only if for
every meet-irreducible open subset~$W$ there is a unique point~$y \in Y$ such
that~$W = Y \setminus \overline{\{y\}}$.
(This is a classical definition of sobriety.)

@@

\begin{exercise}{Complete metric spaces are sober}
A \emph{metric space} is a set~$Y$ together with a relation~$d \subseteq Y
\times Y \times \QQ_{\geq0}$, written~``$d(x,y) \leq q$'', such that the
following axioms are satisfied.
\begin{align*}
  d(x,y) \leq q \wedge q \leq r &\Longrightarrow d(x,y) \leq r \\
  (\forall r \in \QQ_{> q}\+ d(x,y) \leq r) &\Longrightarrow d(x,y) \leq q \\
  &\mathrel{\phantom{\Longrightarrow}} \exists q \in \QQ_{\geq0}\+ d(x,y) \leq q \\
  &\mathrel{\phantom{\Longrightarrow}} d(x,x) \leq 0 \\
  d(x,y) \leq 0 &\Longrightarrow x = y \\
  d(x,y) \leq q &\Longrightarrow d(y,x) \leq q \\
  d(x,y) \leq q \wedge d(y,z) \leq r &\Longrightarrow d(x,z) \leq q+r
\end{align*}
A \emph{Cauchy process} (or \emph{multi-valued modulated Cauchy sequence}) in a metric
space~$Y$ is a map~$\alpha : \QQ_{>0} \to P(Y)$ such that all the sets~$\alpha(q)$
are inhabited and such that~$d(x,x') \leq q + q'$ for all~$q,q' \in \QQ_{>0}$
and~$x \in \alpha(q), x' \in \alpha(q')$. Such a process \emph{converges
to}~$x_0 \in Y$ iff~$d(x,x_0) \leq q$ for all~$q \in \QQ_{>0}$
and~$x \in \alpha(q)$. A metric space~$Y$ is \emph{complete} iff every Cauchy
process in~$Y$ converges.

@@

\begin{exercise}{The covariant approach to locales}%
Let~$A$ be a frame.
\begin{alphlist}[(c)]
\item Show that~$A$ contains arbitrary (set-indexed) meets (though they are not
required to be preserved by frame homomorphisms), by the impredicative
construction
\[ \bigwedge_{i \in I} U_i = \bigvee\{ V \in A \,|\, \text{$V \preceq U_i$ for all~$i \in I$}
\}. \]
\item What are the arbitrary meets in the frame of open subsets of a
topological space?
\item Show that for any frame homomorphism~$\alpha : A \to A'$ there exists a
monotone map~$\beta : A' \to A$ such that~\emph{$\beta$ is right adjoint
to~$\alpha$}, that is such that
\[ \alpha(U) \preceq V \quad\text{iff}\quad U \preceq \beta(V) \]
for all~$U \in A$ and~$V \in A'$.
\end{alphlist}
{\scriptsize\emph{Note.} Based on the observation in part~(c), a
\emph{covariant approach} to locales can be developed, in which locale maps~$X
\to X'$ are defined to be certain kinds of maps~$\OOO(X) \to
\OOO(X')$~\cite{picado-pultr:covariant}.\par}
\end{exercise}

\begin{exercise}{Points as complete prime filters}%
\label{ex:points-as-filters}%
Verify that the points of a locale~$X$ are in canonical one-to-one correspondence
with the completely prime filters of~$\OOO(X)$.
\end{exercise}

\begin{exercise}{Dense opens}%
\label{ex:dense-opens}%
An open~$U$ of a locale is \emph{dense} iff for any open~$V$, $U \wedge V =
\bot$ implies~$U = \bot$.
\begin{alphlist}[(b)]
\item Let~$M$ be a set and endow it with the discrete topology. Show that an
open~$U \in \OOO(L(M))$ is dense iff for any element~$x \in M$, $\neg\neg(x \in
U)$.
\item Let~$\varphi$ be a truth value, hence an open of the one-point
locale~$\pt$. Show that~$\varphi$ is dense in~$\pt$ iff~$\neg\neg\varphi$.
\end{alphlist}
\end{exercise}

\begin{exercise}{Characterizing spatial locales}%
Let~$X$ be a locale. Show that the following conditions are equivalent.
\begin{alphlist}[(c)]
\item For any opens~$U, V \in \OOO(X)$: If~$x \inplus U \Rightarrow y \inplus V$
for all points~$x$ of~$X$, then~$U \preceq V$.
\item For any opens~$U, V \in \OOO(X)$: If~$x \inplus U \Leftrightarrow y \inplus V$
for all points~$x$ of~$X$, then~$U = V$.
\item There is a topological space~$Y$ and a surjective morphism~$L(Y) \to X$
of locales. (A morphism of locales is \emph{surjective} iff its underlying
frame homomorphism is injective.)
\item The canonical morphism~$L(\Pt(X)) \to X$ is an isomorphism.
\end{alphlist}
\end{exercise}

@@

    \textnormal{double rule for existential quantification} \\\smallskip
    \phantom{a}\hfill
    \Axiom$\varphi \ \fCenter\seq{\vec x, y} \psi$
    \doubleLine
    \UnaryInf$\exists y\?Y\+ \varphi\ \fCenter\seq{\vec x} \psi$
    \DisplayProof
    \makebox[0pt]{\shifttext{\tiny ($y$ not free in~$\psi$)}}
    \hfill\phantom{a}
    \bigskip

    \textnormal{mixed rules (unnecessary in presence of the rule for implication)} \\\smallskip
    \vspace{-0.5em}
    \phantom{a}\hfill
    \AxiomC{$\phantom{\seq{\vec x}}$}
    \UnaryInfC{$(\bigvee_i \varphi_i) \wedge \psi \seq{\vec x} \bigvee_i (\varphi_i \wedge \psi)$}
    \DisplayProof
    \hfill
    \AxiomC{$\phantom{\seq{\vec x}}$}
    \UnaryInfC{$(\exists y\?Y\+ \varphi) \wedge \psi \seq{\vec x} \exists y\?Y\+ (\varphi \wedge \psi)$}
    \DisplayProof
    \makebox[0pt]{\shifttext{\tiny ($y$ none of the~$\vec x$)}}
    \hfill\phantom{a}
    \bigskip

@@

    \textnormal{double rule for universal quantification} \\\smallskip
    \phantom{a}\hfill
    \Axiom$\varphi \ \fCenter\seq{\vec x, y} \psi$
    \doubleLine
    \UnaryInf$\varphi\ \fCenter\seq{\vec x} \forall y\?Y\+ \psi$
    \DisplayProof
    \makebox[0pt]{\shifttext{\tiny ($y$ not free in~$\varphi$)}}
    \hfill\phantom{a}
    \bigskip

@@

By adding the axiom
\begin{align*}
  \top &\seq{x:R} (x = 0) \vee (\exists y\?R\+ xy = 1),
\intertext{we describe \emph{geometric fields}, and by additionally adding for
each natural number~$n \geq 1$ the axiom}
  \top &\seq{a_0:R,\ldots,a_{n-1}:R} \exists x\?R\+ x^n + a_{n-1}x^{n-1} +
  \cdots + a_1x + a_0 = 0,
\end{align*}
we describe \emph{algebraically closed geometric fields}. (This field
condition is called ``geometric'' because it can be put as a geometric sequent.
Other, constructively not equivalent conditions such as ``any nonzero element
is invertible'' or ``any nonunit is zero'' are not geometric sequents.)

@@

A variant of this theory is the theory of inhabited objects, which has the
axiom~$\top \vdash \exists x\?X\+ \top$. Its set-based models are precisely the
inhabited sets.

@@

The Lindenbaum algebra is a frame, with finite meets given by~$[\varphi] \wedge
[\psi] = [\varphi \wedge \psi]$ and set-indexed joins given by~$\bigvee_i
[\varphi_i] = [\bigvee_i \varphi_i]$;\footnote{This short description of the
set-indexed joins picks representatives from each equivalence class and is
hence only sensible in the presence of the axiom of choice. In a
constructive metatheory, we should rather write~$\bigvee_{i \in I} M_i =
[\bigvee_{\varphi \in K} \varphi]$, where~$K = \{ \varphi \,|\, \exists i \in
I\+ \varphi \in M_i \}$.} we can regard the Lindenbaum algebra as the free
frame generated by the nullary relation symbols of~$\TT$ modulo the axioms
of~$\TT$. Our interest in the Lindenbaum algebra is because it gives rise to a
locale:
